Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-02-11 multi_body_interactions/papers/theory/paper.tex
renaming referee response file, and adding some misc. revisions

In addition to spectral measurements of many-body interaction
energies, we consider the dynamics of multiply-occupied lattice sites
during spectroscopic interrogation.  While these dynamics do not
provide information about the nature or origin of effective multi-body
interactions, they provide tools and intuition for addressing the
low-lying orbital excitations which are readily accessible in an
experimental setting.  If we initialize all atoms in the $N$-body
ground state with an incoherent mixture of all nuclear spins, then we
prepare the mixed state $\rho_{N,0}=\P_{N,0}/\Tr\P_{N,0}$, where
\begin{align}
  \P_{NX} \equiv \sum_{\abs{\N}=N}\op{\N X}
  \label{eq:P_NX}
\end{align}
is a projector onto the space of the $N$-body orbital states
$\ket{\N X}$.  Interrogating the atoms for a time $t$ by a laser
resonant with the excitation energy $\Delta_{N\pm}$ then gives us the
state
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\Tr\P_{N,0}}
  \sum_{\abs{\N}=N} \exp\p{-it\Omega_{\N\pm}S_{\N\pm}^\x}
  \op{\N,0} \exp\p{it\Omega_{\N\pm}S_{\N\pm}^\x},
\end{align}
where the Rabi frequencies $\Omega_{\N,+},\Omega_{\N,-}$ and
pseudo-spin operators $S_{\N\pm}^\x$ are respectively given in
\eqref{eq:O_N+}, \eqref{eq:O_N-}, and \eqref{eq:S_NX_x}.  Denoting the
eigenstates of $S_{\N\pm}^\x$ by
\begin{align}
  \ket{\N,\S_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} + \ket{\N\pm}},
  &&
  \ket{\N,\A_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} - \ket{\N\pm}},
\end{align}
and defining the identity operator projected to the relevant subspace,
\begin{align}
  \1_{\N\pm} \equiv \op{\N,0} + \op{\N\pm}
  = \op{\N,\S_\pm} + \op{\N,\A_\pm},
\end{align}
we can write the state $\rho_N^{(\pm)}\p{t}$ and excited-state
projectors $\P_{N\pm}$ in the form
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\Tr\P_{N,0}} \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N\pm} + e^{i2t\Omega_{\N\pm}}\op{\N,\S_\pm}{\N,\A_\pm}
    + e^{-i2t\Omega_{\N\pm}}\op{\N,\A_\pm}{\N,\S_\pm}},
  \label{eq:rho_N_pm_t}
\end{align}
and
\begin{align}
  \P_{N\pm} = \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N\pm}-\op{\N,\S_\pm}{\N,\A_\pm}-\op{\N,\A_\pm}{\N,\S_\pm}},
\end{align}
from which it follows that the net excited-state population at time
$t$ is
\begin{align}
  \bk{\P_{N\pm}\p{t}}
  \equiv \Tr\sp{\rho_N^{(\pm)}\p{t}\P_{N\pm}}
  = \f12 - \f12 \bk{\cos\p{2t\Omega_{\N\pm}}}_{\abs{\N}=N},
  \label{eq:P_N_pm_t}
\end{align}
where $\bk{X}_{\abs{\N}=N}\equiv\sum_{\abs{\N}=N}X/\Tr\P_{N,0}$ is an
average of $X$ over all choices of $N$ distinct nuclear spins.

\begin{figure}
  \centering
  \includegraphics{time_table.pdf}
  \caption{\footnotesize Net population of the $N$-body orbital
    excited states $\set{\ket{\N\pm}}$ after interrogation of an
    initial mixed state $\rho_{N,0}=\P_{N,0}/\Tr\P_{N,0}$ for a
    reduced time $\tau_N\equiv t\Omega_0/\sqrt{N}$ (i.e.~with real
    time $t$) by a laser with bare Rabi frequency $\Omega_0$ which is
    resonant with the $N$-body excitation energy $\Delta_{NX}$.  Here
    $\P_{NX}$, defined in \eqref{eq:P_NX}, is a projector onto the
    space of the $N$-body orbital states $\ket{\N X}$.
    \label{fig:time}}
\end{figure}

@@

To understand the periodic collapse and revival of symmetric-state
populations in figure \ref{fig:time}, we observe from \eqref{eq:O_N+}
that the symmetric-state phases in \eqref{eq:rho_N_pm_t} and
\eqref{eq:P_N_pm_t} take the form
\begin{align}
  2t\Omega_{\N,+} = \tau_N \sum_{\mu\in\N} 2\mu
  &&
  \t{with}
  &&
  \tau_N \equiv \f{t\Omega_0}{\sqrt{N}},
  \label{eq:2tO_N+}
\end{align}
where for fermionic atoms with half-integer nuclear spin, $2\mu$ is
always an odd integer, which implies that the sum in \eqref{eq:2tO_N+}
is an integer with the same parity (i.e.~even/odd) as the occupation
number $N$ (i.e.~the number of elements in $\N$).  At reduced times
$\tau_N=n\pi$ with integer $n$, therefore, if the occupation number
$N$ is even then all phases $2t\Omega_{\N,+}$ are integer multiples of
$2\pi$, which leads to a collapse of the excited-state populations as
$\left.\rho_N^{(\pm)}\p{t}\right|_{\tau_N=n\pi} = \rho_N^{(\pm)}\p{0}
= \rho_{N,0}$.  If the occupation $N$ is odd, meanwhile, then the
phases $2t\Omega_{\N,+}$ are all odd (even) integer multiples of $\pi$
for odd (even) $n$.  This alignment of phases implies a complete
population transfer to the excited state
$\rho_{N,-}\equiv\P_{N,-}/\Tr\P_{N,-}$ for odd $n$, and a collapse
back to the orbital ground state $\rho_{N,0}$ for even $n$, precisely
as observed in figure \ref{fig:time}.

@@

We can perform a canonical transformation between $\G$ and $\G_0$
which yields an {\it effective Hamiltonian} $H_{\t{eff}}$ on $\G_0$
that reproduces the spectrum of $H$ on $\G$
\cite{bravyi2011schrieffer}.  Given an eigenbasis
$\set{\ket{\alpha_0}}$ for $H_0$ on $\G_0$ and $\set{\ket\alpha}$ for
$H$ on $\G$, this transformation is implemented by a unitary $U$ for
which $\ket{\alpha_0}=U\ket\alpha$ and $U\to\1$ as $\norm{V}\to0$.
The effective Hamiltonian is then simply
\begin{align}
  H_{\t{eff}} = U H U^\dag.
  \label{eq:H_eff}
\end{align}
The prescription in \eqref{eq:H_eff} for constructing an effective
Hamiltonian is commonly known as a Schieffer-Wolff transformation
\cite{schrieffer1966relation}.  Unitaries $U$ which follow this
prescription are not unique, and different choices of $U$ amount to
different realizations of the Schieffer-Wolff transformation.  In
ref.~\cite{bravyi2011schrieffer}, the authors construct the unique
operator $S$ which generates a {\it direct} or {\it minimal} rotation
$U_{\t{min}}=e^S$ between $\G$ and $\G_0$, and use this construction
to expand \eqref{eq:H_eff} as a perturbative series in $V$.  The
rotation $U_{\t{min}}$ is minimal in the sense that it minimizes the
distance of candiate unitaries $U$ from the identity $\1$ with respect
to the Euclidian operator norm\footnote{The Euclidean operator norm is
  also known as the $L_{2,2}$, Hilbert-Schmidt, or Frobenius norm.}
$\norm{X}_{\t{E}}\equiv\sqrt{\Tr\p{X^\dag X}}$.  This rotation is
determined uniquely by enforcing
\begin{enumerate*}
\item that the generator $S$ is strictly block-off-diagonal with
  respect to $\G_0$ and $\E_0$,
\item that the norm $\norm{S}_{\t{E}}<\pi/2$, and
\item that the block-off-diagonal parts of \eqref{eq:H_eff} are zero.
\end{enumerate*}

>>>>

In addition to spectral measurements of many-body interaction
energies, we consider the dynamics of multiply-occupied lattice sites
during spectroscopic interrogation.  While these dynamics do not
provide information about the nature or origin of effective multi-body
interactions, they provide tools and intuition for addressing the
low-lying orbital excitations which are readily accessible in an
experimental setting.  If we initialize all atoms in the $N$-body
ground state with an incoherent mixture of all nuclear spins, then we
prepare the mixed state $\rho_{N,0}=\P_{N,0}/\tr\P_{N,0}$, where
\begin{align}
  \P_{NX} \equiv \sum_{\abs{\N}=N}\op{\N X}
  \label{eq:P_NX}
\end{align}
is a projector onto the space of the $N$-body orbital states
$\ket{\N X}$.  Interrogating the atoms for a time $t$ by a laser
resonant with the excitation energy $\Delta_{N\pm}$ then gives us the
state
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\tr\P_{N,0}}
  \sum_{\abs{\N}=N} \exp\p{-it\Omega_{\N\pm}S_{\N\pm}^\x}
  \op{\N,0} \exp\p{it\Omega_{\N\pm}S_{\N\pm}^\x},
\end{align}
where the Rabi frequencies $\Omega_{\N,+},\Omega_{\N,-}$ and
pseudo-spin operators $S_{\N\pm}^\x$ are respectively given in
\eqref{eq:O_N+}, \eqref{eq:O_N-}, and \eqref{eq:S_NX_x}.  Denoting the
eigenstates of $S_{\N\pm}^\x$ by
\begin{align}
  \ket{\N,\S_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} + \ket{\N\pm}},
  &&
  \ket{\N,\A_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} - \ket{\N\pm}},
\end{align}
and defining the identity operator projected to the relevant subspace,
\begin{align}
  \1_{\N\pm} \equiv \op{\N,0} + \op{\N\pm}
  = \op{\N,\S_\pm} + \op{\N,\A_\pm},
\end{align}
we can write the state $\rho_N^{(\pm)}\p{t}$ and excited-state
projectors $\P_{N\pm}$ in the form
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\tr\P_{N,0}} \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N\pm} + e^{i2t\Omega_{\N\pm}}\op{\N,\S_\pm}{\N,\A_\pm}
    + e^{-i2t\Omega_{\N\pm}}\op{\N,\A_\pm}{\N,\S_\pm}},
  \label{eq:rho_N_pm_t}
\end{align}
and
\begin{align}
  \P_{N\pm} = \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N\pm}-\op{\N,\S_\pm}{\N,\A_\pm}-\op{\N,\A_\pm}{\N,\S_\pm}},
\end{align}
from which it follows that the net excited-state population at time
$t$ is
\begin{align}
  \bk{\P_{N\pm}\p{t}}
  \equiv \tr\sp{\rho_N^{(\pm)}\p{t}\P_{N\pm}}
  = \f12 - \f12 \bk{\cos\p{2t\Omega_{\N\pm}}}_{\abs{\N}=N},
  \label{eq:P_N_pm_t}
\end{align}
where $\bk{X}_{\abs{\N}=N}\equiv\sum_{\abs{\N}=N}X/\tr\P_{N,0}$ is an
average of $X$ over all choices of $N$ distinct nuclear spins.

\begin{figure}
  \centering
  \includegraphics{time_table.pdf}
  \caption{\footnotesize Net population of the $N$-body orbital
    excited states $\set{\ket{\N\pm}}$ after interrogation of an
    initial mixed state $\rho_{N,0}=\P_{N,0}/\tr\P_{N,0}$ for a
    reduced time $\tau_N\equiv t\Omega_0/\sqrt{N}$ (i.e.~with real
    time $t$) by a laser with bare Rabi frequency $\Omega_0$ which is
    resonant with the $N$-body excitation energy $\Delta_{NX}$.  Here
    $\P_{NX}$, defined in \eqref{eq:P_NX}, is a projector onto the
    space of the $N$-body orbital states $\ket{\N X}$.
    \label{fig:time}}
\end{figure}

@@

To understand the periodic collapse and revival of symmetric-state
populations in figure \ref{fig:time}, we observe from \eqref{eq:O_N+}
that the symmetric-state phases in \eqref{eq:rho_N_pm_t} and
\eqref{eq:P_N_pm_t} take the form
\begin{align}
  2t\Omega_{\N,+} = \tau_N \sum_{\mu\in\N} 2\mu
  &&
  \t{with}
  &&
  \tau_N \equiv \f{t\Omega_0}{\sqrt{N}},
  \label{eq:2tO_N+}
\end{align}
where for fermionic atoms with half-integer nuclear spin, $2\mu$ is
always an odd integer, which implies that the sum in \eqref{eq:2tO_N+}
is an integer with the same parity (i.e.~even/odd) as the occupation
number $N$ (i.e.~the number of elements in $\N$).  At reduced times
$\tau_N=n\pi$ with integer $n$, therefore, if the occupation number
$N$ is even then all phases $2t\Omega_{\N,+}$ are integer multiples of
$2\pi$, which leads to a collapse of the excited-state populations as
$\left.\rho_N^{(\pm)}\p{t}\right|_{\tau_N=n\pi} = \rho_N^{(\pm)}\p{0}
= \rho_{N,0}$.  If the occupation $N$ is odd, meanwhile, then the
phases $2t\Omega_{\N,+}$ are all odd (even) integer multiples of $\pi$
for odd (even) $n$.  This alignment of phases implies a complete
population transfer to the excited state
$\rho_{N,-}\equiv\P_{N,-}/\tr\P_{N,-}$ for odd $n$, and a collapse
back to the orbital ground state $\rho_{N,0}$ for even $n$, precisely
as observed in figure \ref{fig:time}.

@@

We can perform a canonical transformation between $\G$ and $\G_0$
which yields an {\it effective Hamiltonian} $H_{\t{eff}}$ on $\G_0$
that reproduces the spectrum of $H$ on $\G$
\cite{bravyi2011schrieffer}.  Given an eigenbasis
$\set{\ket{\alpha_0}}$ for $H_0$ on $\G_0$ and $\set{\ket\alpha}$ for
$H$ on $\G$, this transformation is implemented by a unitary $U$ for
which $\ket{\alpha_0}=U\ket\alpha$ and $U\to\1$ as $\norm{V}\to0$.
The effective Hamiltonian is then simply
\begin{align}
  H_{\t{eff}} = U H U^\dag.
  \label{eq:H_eff}
\end{align}
The prescription in \eqref{eq:H_eff} for constructing an effective
Hamiltonian is commonly known as a Schieffer-Wolff transformation
\cite{schrieffer1966relation}.  Unitaries $U$ which follow this
prescription are not unique, and different choices of $U$ amount to
different realizations of the Schieffer-Wolff transformation.  In
ref.~\cite{bravyi2011schrieffer}, the authors construct the unique
operator $S$ which generates a {\it direct} or {\it minimal} rotation
$U_{\t{min}}=e^S$ between $\G$ and $\G_0$, and use this construction
to expand \eqref{eq:H_eff} as a perturbative series in $V$.  The
rotation $U_{\t{min}}$ is minimal in the sense that it minimizes the
distance of candiate unitaries $U$ from the identity $\1$ with respect
to the Euclidian operator norm\footnote{The Euclidean operator norm is
  also known as the $L_{2,2}$, Hilbert-Schmidt, or Frobenius norm.}
$\norm{X}_{\t{E}}\equiv\sqrt{\tr\p{X^\dag X}}$.  This rotation is
determined uniquely by enforcing
\begin{enumerate*}
\item that the generator $S$ is strictly block-off-diagonal with
  respect to $\G_0$ and $\E_0$,
\item that the norm $\norm{S}_{\t{E}}<\pi/2$, and
\item that the block-off-diagonal parts of \eqref{eq:H_eff} are zero.
\end{enumerate*}

### 2019-02-11 multi_body_interactions/papers/theory/reviews_NJP/revisions.tex
renaming referee response file, and adding some misc. revisions

Excerpts from the referee reports are written in \blue{blue}, excerpts
from our original manuscript are written in \red{red}, and new text in
the revised version of our manuscript is written in \green{green}.
All page, line, and reference numbers generally refer to those in our
original manuscript, unless explicitly stated otherwise.

@@

  Finally, we have fixed a few discrepancies in our general convention
  that multiple subscripts on a given mathematical symbol should only
  be separated by a comma if at least one of these subscripts has a
  fixed value (as e.g.~in $K^{k\ell}_{mn}$, $K^{\ell,3}_{2,5}$,
  $\c_{\mu n}$, and $\c_{\mu,\g}$).  To this end, in our last appendix
  (page 40, line 17) we have replaced all instances of $U_{M,X}$ and
  $U_{M,X}^{(p)}$ respectively with $U_{MX}$ and $U_{MX}^{(p)}$.

@@

  Both examples are correct due to the fact that
  $G^{\mu q;\nu r}_{\rho s;\sigma t}$ is self-adjoint, but the latter
  ordering convention is used throughout the rest of the manuscript.

@@

>>>>

Excerpts from the referee reports are written in \blue{blue}, excerpts
from our original manuscript are written in \red{red}, and excerpts
from the revised version of our manuscript is written in
\green{green}.  All page, line, equation, and reference numbers
generally refer to those in our original manuscript, unless explicitly
stated otherwise.

@@

  Finally, we have fixed a few discrepancies in our general convention
  that multiple subscripts on a given mathematical symbol should only
  be separated by a comma if at least one of these subscripts has a
  fixed value (as e.g.~in $K^{k\ell}_{mn}$, $K^{\ell,3}_{2,5}$,
  $\c_{\mu n}$, and $\c_{\mu,\g}$).  To this end, we have removed the
  commas on the subscripts ``$M,X$''; ``$N,\pm$''; and ``$\N,\pm$''
  whenever they appear.

@@

  Both examples are ``correct'' due to the fact that
  $G^{\mu q;\nu r}_{\rho s;\sigma t}$ is self-adjoint, but the latter
  ordering convention is used throughout the rest of the manuscript:
  $G^{\mu q;\nu r}_{\rho s;\sigma t}$ is gives the coupling constant
  for destroying two atoms in state $\p{\mu q}+\p{\nu r}$, and
  creating them in $\p{\rho s}+\p{\sigma t}$.

@@

\item For clarity, we have replaced a few instances of the text
  ``\red{single-particle}'' by ``\green{non-interacting}'' when
  referring to motional ground states of non-interacting particles;
  specifically, we have done so on page 9, line 50; on page 13, line
  5; and on page 35, line 26.
